Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability

### 2015-08-21 manuscript.Rmd
Edited according to co-authors comments on revision; suppinfo has own template

\
In ecosystems where water availability limits productivity, we expect an especially large response of *L* to eC~a~ [@Donohue2009]. This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability. These 'water savings' can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~]. Based on the same water-savings principle, we predicted that *L* should increase under eC~a~ in water-limited forests.

@@

\
We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment. The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L* and its components (litter fall and production) in response to water availability. We hypothesize that eC~a~ would result in an increase in *L*, particularly during periods of water limitation. In addition, we expect that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of *L* to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan). Our second objective is to quantify the temporal dynamics of *L* and its components (litter fall and leaf production), in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.). The vegetation consists of a native woodland undisturbed (unmanaged) for ca. 70 years and maintained in this location since the 1880's. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (1881 - 2014, Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). An estimate of potential evapotranspiration (PET) at the site (1350 mm) was from the Global-PET database [1950 - 2000 climatology, http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca. 0.59. In addition, over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating frequent water limitation.

\
At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 - 12.9 m below the soil surface. The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r f(mean(ba$BA))`, SE = `r f(sd(ba$BA)/sqrt(6))` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E. amplifolia* Naudin.  The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as 'rings'). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received aC~a~, and are used as control plots. The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above aC~a~ in full operation mode. However, C~a~ was increased gradually for six months at a rate of ca. 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode and the target eC~a~ was reached in February 2013 (Fig. 1a).

@@

## Terminology
In the following, we use the term leaf area index (*L*) although the canopy and FACE rings contain non-leaf shading elements, i.e. woody stems and infrastructure. This avoids the need to switch between different terminologies when comparing temporal changes in *L* to litter production, and absolute values between different methods. The temporal changes in *L* are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.

## Leaf litter production
In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at random locations (**can I say this?**) at 0.5m above the forest floor. To minimize contamination from trees outside the rings, the baskets were installed at least 2m from the plot edge. Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* by multiplying it with an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14). We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.

## *L* from diffuse transmittance
In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 October 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged at 30-min time steps. At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30-min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca. 5 m above the canopy. For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98. We further trimmed the data by excluding low solar elevation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

\
We then estimated leaf area index (*L*) from $\tau_d$ following Campbell and Norman [-@Campbell2000]. We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where L is *L*, $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for *L* in Eq. 2, using a smooth integration over $\theta$, yielding *L*.

\
To overcome violations in the assumptions underlying Eq. 2, we calibrated *L* against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig. 1). During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm). A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. Cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance. We calibrated Eq. 3,

@@

where *c* is a calibration parameter, $L_{\tau,t=0}$ is *L* at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`). It should be stressed that even after calibration with litter production, *L* still includes stem and infrastructure area (see Terminology). However, this approximately constant area cancels out in the left-hand side of Eq. 3. We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought period (see Fig. S1).

@@

## Data analysis
We used generalized additive models (GAM) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R version `r r_version` [@RCT2015]. A GAM is a semi-parametric fitting method that allows the estimation of smoothed trends in the data, including confidence intervals. In all fitted GAMs, we used a cubic regression spline. For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed *L* dynamics. Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data. Random effects were included to account for the repeated measures structure of the data, except when we analysed average *L* across all rings (Fig. 6).

@@

\
We found no differences in *L* between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig. 2). A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig. 3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on *L* (p > 0.5) (Table 1).

\
Litter production (P~L~) correlated significantly with $\Delta$*L* when the latter was separated into periods where $\Delta$*L* > 0 (net leaf growth) and $\Delta$*L* <  0 (net leaf loss) (Fig. 3). During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where *L* was decreasing (see also Fig. 6). During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta$*L* and litter production (Fig. 4).

@@

\
Next, we related total P~L~ to average *L* (during the litter collection period) across the six FACE rings (Fig. 5a). If leaf lifespan (LL) is constant, P~L~ would be linearly related to *L*. Instead, we found that in eC~a~, P~L~ was lower at a given *L* as compared to aC~a~ (Fig. 6a). We found a significant (p `r pval(LLtest$p.value)`) increase in LL with eC~a~ (Fig. 5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.

\
The dynamics of *L* from the sensor record matched leaf growth events observed in the field quite well (Fig. 6a). All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an observed leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig. 3 and 5), when the soil water content in the top 80 cm increased to ca. > 0.2 m^3^ m^-3^. Two periods of sustained decrease in *L* (red lines in Fig. 6b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

## Response of *L* to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013. We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of eC~a~ on *L*. Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig. 5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur. It could be argued that the change in *L* in response to eC~a~ was too small to be detected. Natural variation in *L* among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a 6.1% in *L* (see Fig. S5). Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to eC~a~.

\
This lack of a response to eC~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons. First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. `r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expect about a 30% increase in *L* under eC~a~. In contrast to other forest FACE sites, the EucFACE is strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015]. Hence, it may be too early to draw general conclusions across FACE sites in terms of *L* response.

@@

\
Despite a lack of response of canopy *L*, we did find a significantly longer LL in eC~a~. We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig. 5). Although LL was increased in eC~a~, this was not reflected in a difference in leaf litter production at any time point (Fig. 4), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought. There is a variety of reports from other FACE sites on the relationship between LL and eC~a~. Norby et al. [-@Norby2003] reported inconsistent effects of eC~a~ on leaf area duration in a deciduous *Liquidambar* plantation (ORNL-FACE). At the same site, Warren et al. [-@Warren2011] found increased senescence in the eC~a~ treatment in response to a combined drought and heat wave. McCarthy et al. [-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* or *Liquidambar* canopies at the DukeFACE, despite enhancement of *L* by eC~a~. Taylor et al. [-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE). Although this indicates the canopy stayed green longer in eC~a~, Gielen et al. [-@Gielen2001] reported faster turnover rate of leaves in eC~a~ at the PopFACE, leading to similar *L* in aC~a~ despite higher productivity. Li et al. [-@Li2000] reported delayed senescence of the oldest leaf cohort in evergreen *Quercus myrtifolia* exposed to eC~a~ in an open-top chamber experiment. Clearly, no consistent picture emerges from these comparisons.

## Litter production and dynamics of *L*
We found a correlation between *L* dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing *L* (Fig. 3). As expected, during periods of decreasing *L*, the change in *L* was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of *L* against litter production. During periods of increasing *L*, litter fall increased sharply (Figs. 4 and 5), indicating replacement of old leaves by young leaves. Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of December 2013. This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months. Similarly, Stone and Bacon [-@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale. Since leaf flushing has been shown to be closely related to soil water availability in eucalypts [Fig. 6; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig. 4). The consequence is that leaf turnover is highly variable in time.

>>>>

\
In ecosystems where water availability limits productivity, we expect a larger response of *L* to eC~a~ [@Donohue2009]. This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability. These 'water savings' can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008]. In line with this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~]. Based on the same water-savings principle, we predicted that *L* increases under eC~a~ in water-limited forests.

@@

\
We estimated *L* from canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to Free-Air CO~2~ Enrichment (FACE), the EucFACE experiment. The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L* and its components (litter fall and production) in response to water availability. We hypothesize that eC~a~ would result in an increase in *L*, particularly during periods of water limitation. In addition, we expect that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of *L* to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan). Our second hypothesis is that the temporal dynamics of *L* and its components (litter fall and leaf production), are related to to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.). The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (1881 - 2014, Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). Estimated potential evapotranspiration (PET) is 1350 mm [1950 - 2000 climatology, http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca. 0.59. In addition, over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating frequent water limitation.

\
At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strongly P-limited [@Crous2015]. Groundwater depth is 12.6 - 12.9 m below the soil surface. The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r f(mean(ba$BA))`, SE = `r f(sd(ba$BA)/sqrt(6))` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E. amplifolia* Naudin.  The understory stratum is dominated by grasses, with lower densities of forbs and occasional shrubs.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as 'rings'). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation [@Lewin2009]. The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received aC~a~, and are used as control plots. The target daytime C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above aC~a~ in full operation mode. However, C~a~ was increased gradually for six months at a rate of ca. 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode and the target eC~a~ was reached in February 2013 (Fig. 1a, [@gimeno2015]).

@@

## Terminology
In the following, we use the term leaf area index (*L*) although the canopy and FACE rings contain non-leaf shading elements, i.e. woody stems and infrastructure. This avoids the need to switch between different terminologies when comparing temporal changes in *L* to litter production, and absolute values between different methods. The temporal changes in *L* are due to leaf area dynamics only, because infrastructure area is constant, and variation in woody stem area can be assumed negligible since no tree mortality has occurred over the two-year study period.

## Leaf litter production
In each ring, eight litter traps were installed at randomly selected locations. The litter traps were 0.1979 m^2^ each with a depth of 0.4 m, and sat at 0.5m above the forest floor. Litter traps were custom-made from plastic baskets coated with a fine nylon mesh with a 200 $/mu$m mesh size. To minimize contamination with litter from trees outside the rings, the baskets were installed at least 2 m from the plot edge. Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* by multiplying it with an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14). We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) on any of the collection dates.

## *L* from diffuse transmittance
In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height. In each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. Data were logged on data loggers (CR3000, Campbell Scientific) at one minutely intervals beginning on 15 October 2012, and automatically uploaded every day to a data management system.  All PAR data were averaged at 30-min time steps. At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30-min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca. 5 m above the forest canopy. For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98. We further trimmed the data by excluding low solar elevation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

\
We then estimated leaf area index (*L*) from $\tau_d$ following Campbell and Norman [-@Campbell2000]. We assumed that leaves had a spherical leaf angle distribution, and that leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for *L* in Eq. 2, using a smooth integration over $\theta$, yielding *L*.

\
To overcome violations in the assumptions underlying Eq. 2, we calibrated *L* against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig. 1). During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred and *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During this period total precipitation was 42mm, the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm). A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. Cumulative *L* leaf litter production (expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance. We calibrated Eq. 3,

@@

where *c* is a calibration parameter, $L_{\tau,t=0}$ and $L_{\tau,t=1}$ are the *L* at the start and the end of the drought period, respectively. We used linear regression without an intercept to estimate *c* in Eq. 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`). It should be stressed that even after calibration with litter production, *L* still includes stem and infrastructure area (see Terminology). However, this approximately constant area cancels out in the left-hand side of Eq. 3. We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought period (see Fig. S1).

@@

## Data analysis
We used generalized additive models (GAM) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R version `r r_version` [@RCT2015]. A GAM is a semi-parametric fitting method that allows the estimation of smoothed trends in the data, including confidence intervals. In all fitted GAMs, we used a cubic regression spline. For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed *L* dynamics. Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data. Random effects were included to account for the repeated measures structure of the data, except when we analysed average *L* across all rings.

@@

\
We found no differences in *L* between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig. 2). A linear mixed-effects model with C~a~, time and pre-treatment basal area (Table 1) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on *L* (p > 0.5) (Table 1).

\
Litter production (P~L~) correlated significantly with $\Delta$*L* when the latter was separated into periods where $\Delta$*L* > 0 (net leaf growth) and $\Delta$*L* <  0 (net leaf loss) (Fig. 3). During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`, p < 0.001), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where *L* was decreasing (see also Fig. 6). During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`, p < 0.001) between $\Delta$*L* and litter production (Fig. 4), indicating that most shedding of old foliage occurred during periods of active leaf growth.

@@

\
Next, we related total P~L~ to average *L* across the six FACE rings (Fig. 5a). If leaf lifespan (LL) is constant, P~L~ would be linearly related to *L*. Instead, we found that in eC~a~, P~L~ was lower at a given *L* as compared to aC~a~ (Fig. 5a). We found a significant (p `r pval(LLtest$p.value)`) increase in LL with eC~a~ (Fig. 5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.

\
The dynamics of *L* from the sensor record matched leaf growth events observed in the field quite well (Fig. 6a). All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an observed leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig. 6b,c), when the soil water content in the top 80 cm increased to ca. > 0.2 m^3^ m^-3^. Two periods of sustained decrease in *L* (red lines in Fig. 6b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

## Response of *L* to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~). We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of eC~a~ on *L*. Because estimated leaf lifespan (LL) averaged `r round(mean(ba$LL),1)` yrs (Fig. 5), the evergreen canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur. It could be argued that the change in *L* in response to eC~a~ was too small to be detected. Natural variation in *L* among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate. Using this method, we estimated that we can detect a difference as little as 6.1% in *L* (see Fig. S5). Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to eC~a~.

\
This lack of a response to eC~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons. First, in a review of all forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. `r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expected about a 30% increase in *L* under eC~a~. In contrast to other forest FACE sites, the EucFACE is strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015]. Hence, it may be too early to draw general conclusions across FACE sites in terms of *L* response.

@@

\
Despite a lack of response of canopy *L* to eC~a~, we did find a significantly longer LL in eC~a~. We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig. 5). Although LL was increased in eC~a~, this was not reflected in a difference in leaf litter production at any time point (Fig. 4), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought. Previous results from other forest FACE studies are not conclusive in this regard. Norby et al. [-@Norby2003] reported inconsistent effects of eC~a~ on leaf area duration in a deciduous *Liquidambar* plantation (ORNL-FACE). At the same site, Warren et al. [-@Warren2011] reported increased senescence in the eC~a~ treatment in response to a combined drought and heat wave. McCarthy et al. [-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* sp. or *Liquidambar* sp. canopies at the DukeFACE, despite enhancement of *L* by eC~a~. Taylor et al. [-@Taylor2008] reported delayed autumn senescence in *Populus* sp. in two FACE experiments (PopFACE and AspenFACE). Although this indicates the canopy stayed green longer in eC~a~, Gielen et al. [-@Gielen2001] reported faster turnover rate of leaves in eC~a~ at the PopFACE, leading to similar *L* in aC~a~ despite higher productivity. Clearly, no consistent picture emerges from these comparisons.

## Litter production and dynamics of *L*
We found a correlation between *L* dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing *L* (Fig. 3). As expected, during periods of decreasing *L*, the change in *L* was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of *L* against litter production. During periods of increasing *L*, litter fall increased sharply (Fig. 3), indicating replacement of old leaves by young leaves. Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of December 2013. This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months. Similarly, Stone and Bacon [-@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale. Since leaf flushing has been shown to be closely related to soil water availability in eucalypts [Fig. 6; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig. 4). Given the high inter-annual variation in rainfall at the EucFACE site, leaf turnover is highly variable in time.

## Continuous measurement of *L*
Studying canopy leaf area index in an evergreen native woodland poses unique challenges. Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because LL is unknown (instead, we used *L* and litter production to estimate LL). We could also not estimate leaf area from allometric relationships with stem diameter or height, since these equations are not available for mature *E. tereticornis*, and would not be valuable anyway since *L* is temporally variable (Fig. 6). We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected (see Fig. S5), indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs. 1, 4, 6). We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to those during periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999]. As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig. S3), because longer path lengths in the winter result in lower overall transmittance. Using diffuse PAR alone avoids having to correct for these path length effects.

## Consequences for phenology models
Current-generation phenology models are heavily biased towards temperate conditions and deciduous rather than evergreen species [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardson2012; @Savoy2015]. In contrast, phenology of evergreen *Eucalyptus* is more opportunistic, and shows rapid response to water availability (Fig. 6). Specht and Brouwer [-@Specht1975] demonstrated for four *Eucalyptus* species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth. Similarly, Pook  [-@Pook1984a] demonstrated occurrence of leaf production during most months in *Corymbia maculata*, except during winter and during dry periods. Thus there seems to be a minimum temperature threshold for *Eucalyptus* leaf growth. Similarly, we also have not observed leaf flushing in winter during the study period, but at our site low temperatures were confounded with water availability as both winters during the study period were relatively dry (Fig. 6).
